---
title: "In-class Activity : Optimal Population with 'Overhead'"
output: html_notebook
author: "Joshua R. Goldstein, Economic Demography Econ/Demog c175"
---

Preliminary congratulations! You successfully navigated this far, opened Rstudio
and found this in-class exercise!


# Instructions:

1. Choose one person in your breakout room to run the RStudio session. If you can't
decide who, do something reverse alphabetical. The others will help this person.

2. Work your way through the exercise together and write down your answers to the
questions on a piece of paper, e.g., "Q1: A".

3. Have fun. Don't stress if you don't understand everything. 

4. When we regroup (if we have time today), we'll go through the answers and discuss

# Part 1: Diminishing Marginal Returns

The first part of this exercise will be to look at the case of
marginal returns that are diminishing from the beginning (rather
than rising and then falling in many of Sauvy's examples)

Here's our set up: click the green arrow to run the whole chunk. Or run it line
by line using Command-Enter (on a mac).

```{r}
## (1) population sizes from 1 to 50
N = 1:50
print(N)
## (2) marginal product that is declining (in this case exponentially at rate "k")
k = .03
mp <- exp(-k * N)
print(mp)
## (3) total product, using the fancy cumsum() function, try cumsum(1:3) and you should get 1, 3, 6
tp = cumsum(mp)
## (4) average product
ap = tp/N
```

Let's look at what we've done using plots

```{r}
par(mfrow = c(2,2)) ## this creates a 2x2 grid of sub-plots
plot(x = N, y = mp, main = "Marginal product")
plot(x = N, y = tp, main = "Total Product")
plot(x = N, y = ap, main = "Average Product")
```


Q1: is marginal product really decreasing?
A. Yes
B. No
Hint: You should be able to see in the figure.

Q2: For population sizes bigger than one, what is the relationship
    between marginal product and average product?
A. They are equal
B. MP is bigger
C. AP is bigger
Hint: If MP is going down, as we go from 1 to 2 people, say from 1
to .8, what will AP be for 2 people?
(Note: Your answer should be true when MP is declining, but it will
not hold generally, as we saw in the slides and in the reading.)

Q3. What population size maximizes average product?
A. 1
B. 50
C. 10
D. about 15, but hard to see exactly on graph

## Part 2: Overhead

Now let's consider the effect of fixed costs or "overhead", say due
to a population paying "tribute" to another population, or
maintaining a certain size military, or maintaining infrastructure.

Let's consider the case wheen fixed cost = 2.

```{r}
overhead = 2
tp.prime = tp - overhead ## we substract overhead to get Total Product available for population members
ap.prime = tp.prime/N ## and recalculate Average Product

## now we plot the original (without overhead numbers) together with the
## new numbers (with overhead), which end in ".prime"
par(mfrow = c(2,2))
plot(x = N, y = mp, ylim = c(0, 1))  ## ylim here shows us the plot along the y-axis from 0 to 1
title("Marginal product")
plot(N, tp, ylim = c(0, max(tp))) ## here ylim shows us the plot from 0 to the maximum value
title("Total product, \n solid line shows with overhead")
lines(N, tp.prime) ## KEY: lines() adds a line to existing plot, in this case for tp.prime
plot(N, ap, ylim = c(0, 1)) ## and now same for ap
lines(N, ap.prime)
title("Average product, \n solid line shows with overhead")
```


Q3: What is tp.prime when N = 1
A. Less than zero
B. Negative
C. A bit awkward to define
D. All of the above
Hint: Not a really clear answer here, so I would choose "D"

Q4: What population size maximizes the "economic optimum" defined by Sauvy?
A. 1
B. 50
C. 10
D. about 15, but hard to see exactly on graph
Hint: The economic optimum maximizes average product


Let's get R to tell us what the maximum. Don't worry about the
programming. Just run the code and see the result

```{r}
par(mfrow = c(2,2))
plot(x = N, y = mp, ylim = c(0, 1))  ## ylim here shows us the plot along the y-axis from 0 to 1
title("Marginal product")
plot(N, tp, ylim = c(0, max(tp))) ## here ylim shows us the plot from 0 to the maximum value
title("Total product, \n solid line shows with overhead")
lines(N, tp.prime) ## KEY: lines() adds a line to existing plot, in this case for tp.prime
plot(N, ap, ylim = c(0, 1)) ## and now same for ap
lines(N, ap.prime)
title("Average product, \n solid line shows with overhead")
## add a line at the max with a label ont he axis
w = which.max(ap.prime)
abline(v = N[w], lty = 2, col = "red")
axis(side = 1, at = N[w], col = "red", col.axis = "red", font = 4)
```
Does the result match your answer to Q4?


Q5. What would happen if to economic optimum if we _increased_ overhead? Discuss in your group.

Q6. What would be the impact of overhead on the power optimum? Discuss in your group.


Congratulations! You've finished our first in-class lab! (If you didn't get this far
or if you had trouble with Rstudio or anything else, do lab_01_world.Rmd, which offers
more explanation and hand-holding for the computing, and then come back and see if 
you can finish the in-class lab.)
